\chapter{Background}
\label{chap:background}

% Reading guide: each section establishes one conceptual prerequisite and ends
% with an open question or unresolved tension. The gaps identified here are
% resolved in the Methods chapter.

\section{Imagery Rehearsal Therapy}
\label{sec:bg-irt}
% QUESTION THIS SECTION ANSWERS: What is IRT and why is rescripting the hard part?
%
% Five-stage protocol: recording, rewriting, summary, rehearsal, final
%   Each stage has defined purpose and bounded appropriate therapist behaviours
% Rescripting: patient rewrites the nightmare to reduce distress and restore agency
%   This is the cognitive core -- the phase where the therapist must make active
%   strategic decisions (not just listen or summarise)
% Treatment fidelity instruments (ENACT, NIH BCC): define adherence gradations
%   absent / partial / implemented -- the clinical standard for measuring whether
%   a therapist delivers an intervention correctly
% Clinical evidence base for IRT effectiveness
%
% TENSION LEFT OPEN: Fidelity assessment exists for human therapists delivering
%   manualized protocols. No equivalent exists for LLM-based agents -- the
%   evaluation instruments assume a human clinician.
%
% [Example] Nightmare excerpt -> rescripted version with strategy labels overlaid
%   (cognitive_reframe, confrontation) -- what a valid rescripting move looks like

Nightmare disorder affects approximately 4--5\% of the adult population and
shows strong comorbidity with \gls{ptsd} and depression
\parencite{34M7Y3F8}. For patients with recurrent nightmares, the distress
extends beyond the dream itself: sleep avoidance, daytime fatigue, and
heightened anxiety create a cycle that standard sleep hygiene cannot break.
\gls{irt} is the recommended treatment for this condition
\parencite{34M7Y3F8, FGA5H3ZD}.

\gls{irt} follows a structured five-stage protocol. In the \emph{recording}
stage, the patient describes the nightmare in detail. In the \emph{rewriting}
stage, the patient and therapist collaboratively transform the nightmare
narrative to reduce distress and increase a sense of mastery. The remaining
stages (\emph{summary}, \emph{rehearsal}, and \emph{final}) consolidate the
new narrative and prepare the patient to rehearse it before sleep.
\textcite{FGA5H3ZD} describe the rewriting instruction as deliberately
open-ended: patients are told to ``change the nightmare any way you wish.''

The rewriting stage is the cognitive core of the protocol. Unlike recording
(which requires listening) or rehearsal (which requires repetition), rewriting
demands active therapeutic strategy decisions. The therapist must guide the
patient toward a specific rescripting mechanism: confronting the threat,
introducing protective elements, reframing the narrative, or transforming the
dreamer's role within it. These choices are not arbitrary. Clinical research
has identified distinct rescripting strategies with different prevalence rates
in successful treatment. \textcite{25UR8BHJ} coded five categories from
combat veteran rescripts: alternative endings (58\%), positive image insertion
(23\%), threat transformation (13\%), reality markers (10\%), and distancing
(8\%). \textcite{62ZMHBC3} proposed a complementary framework through the
Multidimensional Mastery Scale, which classifies rescripting outcomes along
physical, social, environmental, emotional, mystical, and avoidance
dimensions.

Treatment fidelity instruments provide the clinical standard for measuring
whether a therapist delivers an intervention correctly. The \gls{irt}
literature draws on two established frameworks: the ENACT rating
scale \parencite{GYNNP2CC} and the NIH BCC recommendations
\parencite{ITT882KN}. Both grade therapist adherence on a ternary scale:
\emph{absent}, \emph{partial}, or \emph{implemented}. This gradation
captures clinically meaningful differences. A therapist who briefly mentions
a rescripting strategy without developing it (partial) delivers a different
intervention from one who fully guides the patient through it (implemented).

This ternary framework provides a scoring foundation that transfers directly
to computational evaluation (see \autoref{sec:meth-m3}). However, fidelity
assessment as it stands assumes a human clinician. No equivalent instrument
exists for \gls{llm}-based agents, where the therapist is a probabilistic
text generator rather than a trained professional.

% [Example] Nightmare excerpt -> rescripted version with strategy labels
%   overlaid (cognitive_reframe, confrontation)


\section{Treatment Fidelity in LLM-Based Interventions}
\label{sec:bg-llm-healthcare}
% QUESTION THIS SECTION ANSWERS: Why is therapeutic AI different from a chatbot?
%
% Scope: protocol adherence in manualized interventions, not the full
%   LLM-healthcare landscape. General clinical AI is referenced only where
%   it establishes failure modes relevant to stability measurement.
% The general vs. protocol distinction -- conceptual grounding:
%   Treatment fidelity research (Mowbray et al., TYX7W5ZM; NIH BCC, ITT882KN)
%   formalises what it means for a therapist to deliver an intervention correctly:
%   adherence (doing the prescribed things) and competence (doing them well)
%   Fidelity is only a meaningful concept when there is a protocol to adhere to --
%   it is undefined for a general assistant
%   For rule-based chatbots, adherence is built in by construction -- responses
%   are pre-authored and reviewed once
%   For LLM-based protocol agents, adherence becomes probabilistic -- it must
%   be measured across runs, not assumed from a single review
% Capability vs. reliability -- the core gap:
%   A model can produce high-quality individual responses while being clinically
%   unreliable across sessions -- capability benchmarks cannot detect this
% General LLM failure modes (hallucination, stage confusion, persona
%   inconsistency) exist but are outside this study's scope. This study
%   focuses on the probabilistic nature of LLM outputs creating drift
%   in identical therapeutic contexts.
% Regulatory framing: EU AI Act high-risk classification; Medical Device
%   Regulation auditability and documentation requirements
%   Data sovereignty as a legal constraint pushing toward EU-resident models
%
% TENSION LEFT OPEN: Stability is a precondition for clinical deployment, but
%   no existing evaluation framework measures it for protocol-driven LLM agents.
%   Prior therapeutic AI evaluation is largely qualitative, post-hoc, or
%   single-session -- it cannot detect cross-run inconsistency.
%
% [FIG] EU AI Act risk tier diagram -- where autonomous therapeutic AI sits

The deployment of \glspl{llm} in mental health has expanded from empathy
tools and symptom screeners to protocol-driven therapeutic agents.
\textcite{DTKNEPHQ} survey this trajectory and argue that \glspl{llm} could
fundamentally reshape behavioural healthcare delivery, from diagnostic support
to full psychotherapeutic interventions. Early systems such as \modelname{Woebot}
\parencite{YE8KQTNH} delivered \gls{cbt} through rule-based dialogue trees
where every response was pre-authored and reviewed before deployment. For
these systems, treatment fidelity is a design property: the chatbot cannot
deviate from its script.

\gls{llm}-based agents operate differently. Each response is a fresh sample
from the model's output distribution. A single model can produce
high-quality, protocol-adherent responses on one run and strategically
inconsistent ones on the next. This distinction is central to the treatment
fidelity literature. \textcite{TYX7W5ZM} formalise fidelity into two
components: \emph{adherence} (doing the prescribed things) and
\emph{competence} (doing them well). For rule-based chatbots, adherence is
built in by construction. For \gls{llm}-based protocol agents, adherence
becomes probabilistic and must be measured across runs rather than assumed
from a single review.

\gls{llm}-based agents also face general failure modes such as hallucination,
stage confusion, and persona inconsistency. These are well-documented
concerns for clinical AI but fall outside the scope of this study. The
present work focuses on a narrower problem: the probabilistic nature of
\gls{llm} outputs creates drift even in identical therapeutic contexts. When
the same patient presents the same nightmare, stochastic sampling can produce
different strategy selections, different therapeutic responses, and
mismatches between declared and executed strategies. Each of these failure
modes corresponds to a distinct layer of the generation process and requires
a targeted evaluation method. Strategy drift maps to plan-level measurement
(\autoref{sec:meth-m1}), output variability to semantic comparison
(\autoref{sec:meth-m2}), and plan-output mismatch to alignment verification
(\autoref{sec:meth-m3}).

Regulatory frameworks reinforce the need for stability measurement. The
\gls{eu} AI Act classifies autonomous therapeutic AI as high-risk, imposing
requirements for auditability, documentation, and risk management
\parencite{X5TYM56Q, A823HQRV}. The Medical Device Regulation adds further
constraints on software used in clinical contexts. For \gls{eu}-based
deployment scenarios, data sovereignty requirements push toward self-hosted,
\gls{eu}-resident models \parencite{2QMKF3RM}, constraining the model
selection space (see \autoref{sec:meth-models}).

Despite these requirements, stability remains unmeasured in practice. Prior
therapeutic AI evaluation is largely qualitative, post-hoc, or
single-session. \textcite{B97RI7GA} represent the closest prior work with
their BOLT framework, which assesses \gls{llm} therapist behaviour across
13 behavioural codes drawn from motivational interviewing. BOLT asks whether
the model does the right thing. The present study asks a different question:
whether the model does the \emph{same} thing across repeated runs. No
existing framework measures this cross-run consistency for protocol-driven
\gls{llm} agents.

% [FIG] EU AI Act risk tier diagram


\section{Stochastic Behaviour in Language Models}
\label{sec:bg-stochastic}
% QUESTION THIS SECTION ANSWERS: Where does instability come from mechanically?
%
% Temperature sampling: at T>0, token selection is probabilistic -- small
%   per-token variance compounds across a full therapeutic response
% Autoregressive generation: each token conditions on all previous tokens --
%   a single divergent token can cascade into a completely different strategy
% Clinical consequence: the same patient on two different days may receive
%   strategically inconsistent care -- this is not benign noise in a
%   manualized protocol
% Prior stability work in NLP:
%   Targets factual consistency, output diversity, or benchmark reproducibility
%   Not therapeutic strategy stability or protocol adherence
%
% TENSION LEFT OPEN: The mechanical source of instability is well understood,
%   but its impact on structured therapeutic decision-making has not been
%   quantified. How much does stochastic sampling cost in clinical consistency?
%
% [FIG] Sampling diagram -- same frozen history at T=0.0 vs. T=0.7 ->
%   diverging strategy distributions

At temperatures above zero, \gls{llm} token selection is probabilistic.
The softmax distribution over the vocabulary is divided by the temperature
parameter before sampling, so higher temperatures flatten the distribution
and lower temperatures sharpen it toward the most likely token. At $T = 0$,
the model selects the highest-probability token deterministically (greedy
decoding). At $T > 0$, tokens are sampled from the resulting distribution,
introducing variance into the output.

This per-token variance compounds through autoregressive generation. Each
token is conditioned on all preceding tokens, so a single divergent token
early in a response can cascade into a completely different therapeutic
strategy. A model that selects ``confront'' rather than ``imagine'' at one
token position may generate an entirely different rescripting approach in the
subsequent response. This \emph{autoregressive cascading} means that the
clinical impact of stochastic sampling is not proportional to the per-token
entropy but can amplify through the response.

The clinical consequence is concrete: the same patient presenting the same
nightmare on two different days may receive strategically inconsistent care.
In a general-purpose assistant, this variability is benign or even desirable.
In a manualized therapeutic protocol, where the therapist is expected to
deliver consistent, evidence-based interventions, it represents a potential
clinical failure.

Prior work on \gls{llm} stability targets different concerns. Factual
consistency studies measure whether models produce the same answers to
knowledge questions across runs. Output diversity research quantifies
lexical and semantic variation in creative generation. Benchmark
reproducibility work examines whether evaluation scores are stable across
random seeds.
% TODO: add specific citations for factual consistency, output diversity,
%   and benchmark reproducibility studies once identified
None of these lines of research address therapeutic strategy stability:
whether a model makes the same clinical decisions when presented with
identical patient contexts.

The mechanical source of instability is well understood. What remains
unquantified is its impact on structured therapeutic decision-making. How
much does temperature-driven variance cost in clinical consistency? This
question motivates the temperature conditions in the experimental design
(\autoref{sec:meth-conditions}), where $T = 0.0$ serves as a deterministic
upper bound on stability and $T = 0.7$ represents realistic clinical
deployment.

% [FIG] Sampling diagram -- same frozen history at T=0.0 vs. T=0.7


\section{Evaluation Metrics for Text Generation}
\label{sec:bg-metrics}
% QUESTION THIS SECTION ANSWERS: How do people measure text similarity, and
%   why is it hard for therapy?
%
% Surface metrics (BLEU, ROUGE): token overlap
%   Therapeutically equivalent paraphrases score near zero
%   Lexically similar but strategically divergent responses score high
%   Ruled out: therapeutic paraphrasing is the norm; surface metrics
%   penalise valid variation
% BERTScore (Zhang et al., 2020): contextual embedding similarity
%   Meaning equivalence independent of surface form
%   Multiple embedding model options with different strengths
%   Highest human correlation achieved with NLI-fine-tuned models
%   Open question: how well does this transfer to therapeutic language?
% LLM-as-judge paradigm:
%   Evaluates criteria no scalar metric can capture
%   (e.g. "does this response implement a confrontation strategy?")
%   Known biases: self-preference, position bias, verbosity bias
%   Mitigation strategies in the literature: cross-model judging,
%   deterministic decoding, structured rubrics
%
% TENSION LEFT OPEN: Embedding similarity captures semantic equivalence but
%   cannot assess strategic intent. LLM judges can assess intent but introduce
%   their own reliability concerns. No single metric covers what clinical
%   stability requires -- a multi-level approach is needed, but none exists
%   for therapeutic AI.
%
% [FIG] ROUGE vs. BERTScore on a matched therapeutic sentence pair --
%   where surface overlap fails and why embedding similarity is needed

Measuring consistency between text outputs requires a similarity metric.
The choice of metric determines what kind of consistency is captured, and
standard options fail in different ways for therapeutic language.

Surface-level metrics such as BLEU and ROUGE measure token overlap between
texts. In therapeutic contexts, these metrics produce misleading results in
both directions. Two responses that use identical words to describe different
strategies score high on surface overlap despite being clinically divergent.
Conversely, two responses that implement the same strategy through different
phrasing score low despite being therapeutically equivalent. Therapeutic
paraphrasing is the norm rather than the exception: a competent therapist
will naturally vary their language while maintaining the same strategic
intent. Surface metrics penalise this valid variation.

BERTScore \parencite{PUYCKUYY} addresses this limitation by computing similarity in contextual
embedding space rather than token space. Each token in the candidate text is
matched to the most similar token in the reference text using cosine
similarity over contextualised embeddings, producing precision, recall, and
F1 scores. The metric captures semantic equivalence independent of surface
form: two sentences that mean the same thing but use different words receive
high scores.

The choice of embedding model affects BERTScore's behaviour. Among more than
130 available models, those fine-tuned on \gls{nli} tasks achieve the
highest correlation with human judgements of semantic equivalence.
\modelname{DeBERTa-XLarge-MNLI} \parencite{8MQC2H2V} achieves the highest WMT16 human
correlation ($r = 0.778$) and is selected for Method~2 in this study
(\autoref{sec:meth-m2}). Its disentangled attention mechanism handles the
surface-form variation typical of therapeutic language, where valid
paraphrasing is expected.

BERTScore captures semantic equivalence but cannot assess strategic intent.
A metric that reports high similarity between two responses does not reveal
whether both responses implement the same therapeutic strategy. The
\gls{llm}-as-judge paradigm addresses this gap. By instructing a separate
\gls{llm} to evaluate whether a response implements a declared strategy,
it becomes possible to assess criteria that no scalar metric can capture:
``does this response implement a confrontation strategy?'' requires
understanding both the strategy definition and the response content.

\gls{llm}-as-judge approaches carry known biases. Self-preference bias
causes models to rate outputs from their own family more favourably.
Position bias leads to higher scores for responses presented first.
Verbosity bias rewards longer responses regardless of quality
\parencite{LVLA9LZU}. These biases can be mitigated through cross-model
judging (the judge evaluates a different model family), structured rubrics,
and explicit justification requirements. The mitigation strategy adopted in
this study is detailed in \autoref{sec:meth-m3}.

No single metric covers what clinical stability requires. Embedding
similarity captures semantic equivalence but misses strategic intent.
\gls{llm} judges can assess intent but introduce their own reliability
concerns. Surface metrics capture neither. A multi-level approach is needed,
but no such framework exists for therapeutic AI evaluation. This gap
motivates the three-method design presented in \autoref{chap:methods}.

% NOTE: Chain-of-Thought and Plan Faithfulness content folded into
% Section 3.4 (The Plan Mechanism) where the declared-intent resolution
% is introduced. Citations: Wei et al. 2022, Turpin et al. 2023,
% Lanham et al. 2023.
