\chapter{Methods}
\label{chap:methods}

\section{System Overview and Design Rationale}
\label{sec:meth-overview}
% Two decoupled stacks: generation (dialogue creation) and evaluation
%   (stability measurement)
%   Why decoupled? Evaluation must not be contaminated by generation-side
%   randomness -- reproducible entry points require frozen, pre-generated histories
% End-to-end data flow:
%   vignettes -> generation -> frozen histories -> evaluation -> aggregation
% [FIG] Full pipeline diagram -- logical data flow across all stages, stack
%   boundary marked, example artefact names at each step

The evaluation system separates dialogue creation from stability measurement
through two decoupled stacks. The \emph{generation stack} produces synthetic
therapy sessions by running a multi-agent dialogue loop. The
\emph{evaluation stack} measures the stability of therapeutic responses by
presenting frozen conversation histories to the model under test and
collecting multiple independent outputs.

This separation is a deliberate design choice. If the evaluation stack
generated its own dialogue context, any measured instability could reflect
upstream randomness in the patient or routing agents rather than genuine
variation in the therapist model. By freezing the conversation history before
evaluation begins, the system ensures that every trial receives identical
input. Any observed inconsistency is attributable solely to the model under
test.

The end-to-end data flow proceeds in five stages: patient vignettes feed the
generation stack, which produces complete therapy sessions. These sessions
are sliced at rewriting-turn boundaries to create frozen histories. The
evaluation stack presents each frozen history to the therapist model across
multiple independent trials, collecting both a strategy declaration and a
therapeutic response per trial. Finally, an aggregation pipeline aligns all
metrics across conditions for comparative analysis.

% [FIG] Full pipeline diagram -- two-stack data flow with isolation boundary


\section{Dialogue Generation}
\label{sec:meth-dialogue}
% Three-agent loop: Patient (BDI-profiled vignette), Router (stage classifier),
%   Therapist (stage-appropriate response)
%   Why three agents? Mirrors clinical interaction structure; separates
%   routing logic from therapeutic content generation
% Six patient vignettes: anxious, avoidant, cooperative, resistant, skeptic, trauma
%   Cover the range of clinical presentation difficulty expected in IRT deployment
%   Vignette design informed by prior synthetic patient work (Roleplay-doh??, check ref).
%   Six profiles sample the difficulty range but do not claim exhaustive coverage (7.2).
% Full five-stage IRT traversal produces naturalistic conversation histories
%   used as frozen evaluation entry points
% [FIG] Three-agent loop -- single turn cycle with message passing
% [Example] Resistant vignette excerpt -- BDI profile and opening patient turn

The generation stack orchestrates a three-agent loop that produces
naturalistic \gls{irt} sessions. Each turn proceeds through three agents in
sequence:

\begin{enumerate}
  \item The \textbf{patient agent} simulates a clinical presentation based on
    a \gls{bdi}-profiled vignette. Each vignette defines the patient's
    nightmare content, emotional presentation style, and engagement pattern.
  \item The \textbf{router agent} classifies the current \gls{irt} stage
    from the conversation history, determining whether the session is in the
    recording, rewriting, summary, rehearsal, or final phase.
  \item The \textbf{therapist agent} generates a stage-appropriate
    therapeutic response, guided by the router's classification and a
    stage-specific system prompt.
\end{enumerate}

This three-agent architecture separates clinical reasoning from stage
classification and patient simulation, matching the deployment architecture
of reDreamAI. The router's stage classification controls which system prompt
the therapist receives, ensuring that responses are appropriate to the
current protocol phase.

Six patient vignettes cover the range of clinical presentation difficulty
expected in \gls{irt} deployment: \emph{anxious}, \emph{avoidant},
\emph{cooperative}, \emph{resistant}, \emph{skeptic}, and \emph{trauma}.
These profiles range from a cooperative patient who engages readily with
rescripting instructions to a trauma patient whose nightmare content
triggers intense emotional responses during the session. The vignette design
draws on prior synthetic patient work. \textcite{Z22JYAMU} demonstrate that
expert-defined behavioural principles produce clinically plausible simulated
patients. \textcite{CQ94GEYA} provide cognitive conceptualisation diagrams
and conversational styles (direct, resistant, expressive, minimal, deviating,
agreeable) that parallel the six archetypes used here. Six profiles sample
the difficulty range but do not claim exhaustive coverage of patient
variability (see \autoref{sec:concl-limitations}).

The generation stack traverses the full five-stage \gls{irt} protocol,
producing complete therapy sessions that are then sliced to create
evaluation entry points.

% [FIG] Three-agent loop -- single turn cycle with message passing
% [Example] Resistant vignette excerpt -- BDI profile and opening turn


\section{Frozen History Design}
\label{sec:meth-frozen}
% Deterministic evaluation entry points: identical context for every trial
% Conversations sliced at rewriting-turn boundaries
%   Why frozen? Any measured inconsistency is attributable solely to the
%   model under evaluation -- not to upstream context differences
% Slice depth pilot: 5 depths tested on Mistral Large 3 to select evaluation point
%   Slice 1: introduction to rescripting, moderate variability
%   Slice 2: lowest consistency -- strategy not yet anchored (conservative choice)
%   Slices 3-5: strategy locks in, consistency rises to ceiling
% Main experiment runs at slice 2 (mid-rescripting) -- conservative lower bound
% [FIG] Slicing diagram -- P/T turn sequence with cut points; which turns are
%   in-context vs. excluded per slice

Frozen conversation histories serve as deterministic evaluation entry
points. Each vignette's complete therapy session is sliced at
rewriting-turn boundaries, where each slice ends after a complete
therapist-patient exchange. Every trial, across all models and
temperatures, receives the identical frozen history as input. This design
eliminates context variation as a confound: if two trials produce different
therapeutic responses, the difference is attributable to the model's
stochastic sampling, not to differences in the input conversation.

A preliminary analysis tested \modelname{Mistral Large~3} at five slice
depths across all six vignettes at $T = 0.075$ and $T = 0.15$. Jaccard
consistency was lowest at the second slice (mean $J = 0.77$ at $T = 0.075$)
and rose to near-ceiling values from the third slice onward ($J > 0.93$).
The main experiment evaluates at the second rewriting-turn boundary
(\techil{slice\_2}), making the measured stability a conservative estimate.
Slice depth effects are reported in \autoref{sec:res-secondary}.

% [FIG] Slicing diagram -- P/T turn sequence with slice 2 marked


\section{Evaluation Stack}
\label{sec:meth-eval}
% Router bypassed -- rescripting prompt injected directly (stage known a priori)
%   Why bypass? Isolates the therapist model from stage-classification error
% Fused CoT generation: <plan> block + response in a single model call
%   <plan> block declares 1-2 strategies before the therapeutic response
%   Plan conditions the response tokens (Chain-of-Thought style)
%   Why fused? Plan must condition the response tokens -- a separate call
%   would decouple intent from output
% 20 independent trials per condition sample the stochastic output distribution
% [FIG] Evaluation stack -- frozen history -> injected prompt -> fused generation
% [Example] Raw model output: <plan> block and response side by side

The evaluation stack bypasses the router and injects the rescripting prompt
directly into the therapist model. Because the \gls{irt} stage is known a
priori (all frozen histories are sliced within the rewriting phase), routing
is unnecessary and would introduce an additional source of variation.
Bypassing it isolates the therapist model from stage-classification error.

Each evaluation trial uses \emph{fused generation}: the model produces a
\techil{<plan>} block declaring one to two therapeutic strategies, followed
by the therapeutic response, in a single call. The plan tokens condition the
response tokens in a \gls{cot}-style generation, so the response is
directly conditioned on the declared strategy. This fused approach enables
the alignment analysis in Method~3 (\autoref{sec:meth-m3}): because plan
and response are generated in a single pass, any mismatch between declared
intent and executed strategy reflects the model's own inconsistency rather
than a decoupling artefact.

Twenty independent trials are collected per condition, yielding
$\binom{20}{2} = 190$ pairwise comparisons. This trial count balances
statistical power with compute cost: 190 pairs provide robust coverage
of the pairwise similarity distribution while keeping the total experiment
within budget.

% [FIG] Evaluation stack -- frozen history, injected prompt, fused output
% [Example] Raw model output: <plan> block and response side by side


\section{The Plan Mechanism}
\label{sec:meth-plan}
% The CoT faithfulness problem (folded from Background):
%   CoT prompting (Wei et al., 2022) improves output structure, but
%   faithfulness literature (Turpin et al., 2023; Lanham et al., 2023) shows
%   that CoT explanations often do not reflect actual computation. If a model
%   declares its strategy before responding, that declaration may be post-hoc
%   rationalization rather than a window into its decision-making.
%
% Resolution: <plan> block declares 1-2 IRT strategies before the response
% Framed as declared intent -- a structured output, not a reasoning trace
%   Why? Sidesteps the CoT faithfulness problem: the plan is not claimed to
%   be a window into model cognition. It is a measurable output in its own right.
%   If strategy declarations vary across runs, clinical responses will vary --
%   this is true regardless of whether declarations reflect internal reasoning
% Structured measurement independent of CoT faithfulness debate
% Enables three independent uses:
%   1. Strategy-level consistency scoring (Method 1)
%   2. Plan-response alignment verification (Method 3)
%   3. Human-readable clinical oversight
% Fixed IRT strategy taxonomy constrains declarations to the therapeutic domain --
%   prevents arbitrary strategies that cannot be clinically evaluated
% [FIG] Plan mechanism -- <plan> block feeding into three downstream uses
%   with declared-intent framing labelled

The \techil{<plan>} block raises a methodological question. \gls{cot}
prompting \parencite{KE8ZHW9V} improves output structure by asking models
to show their reasoning before answering. However, the faithfulness
literature demonstrates that \gls{cot} explanations often do not reflect
the model's actual computation. \textcite{TMCBKEVQ} show that biasing cues
in the prompt alter model behaviour without appearing in the \gls{cot}
trace. \textcite{W4K3ARI5} find that early \gls{cot} tokens can be
truncated or replaced without affecting the final answer, suggesting that
the reasoning trace is reconstructed post-hoc rather than used for
computation.

If a model declares its therapeutic strategy before responding, that
declaration may be a post-hoc rationalisation rather than a window into its
decision-making process. This study resolves the tension by framing the
\techil{<plan>} block as \emph{declared intent} rather than a reasoning
trace. The plan captures what the model commits to, not what it internally
computes. Whether or not the declaration reflects genuine reasoning, a
practical consequence holds: if strategy declarations vary across runs,
clinical responses will also vary. Variable declared intent predicts
variable therapeutic behaviour, which is the clinically relevant property.

This framing sidesteps the faithfulness debate entirely while preserving
three concrete uses. First, the declared strategies enable strategy-level
consistency scoring (Method~1, \autoref{sec:meth-m1}). Second, comparison
between declared strategies and the executed response enables alignment
verification (Method~3, \autoref{sec:meth-m3}). Third, the plan provides
a human-readable summary of the model's intended approach, supporting
clinical oversight in deployment.

% [FIG] Plan mechanism -- <plan> block feeding into three downstream uses


\section{Strategy Taxonomy Development}
\label{sec:meth-taxonomy}
% Why fixed? Enables discrete, comparable strategy sets -- free-text plans
%   cannot be aggregated into Jaccard similarity scores
% Iterative revision driven by observed distribution failures:
%   v1 (8 categories): 87.7% mass on empowerment/mastery -> unusable skew
%   v2 (merge to agency): 100% dominance -> goal-mechanism conflation
%   v3 (mechanism-specific split): confrontation (external action) vs.
%     self_empowerment (internal transformation)
% Final 6 categories: confrontation, self_empowerment, safety, cognitive_reframe,
%   social_support, sensory_modulation
% Prompt-level steering encourages diversity across categories
% [GRAPH] Strategy frequency per taxonomy version (v1 -> v3) -- shows skew
%   collapse and final distribution
% [FIG] 6-category taxonomy card -- each category, definition, example phrase

A fixed taxonomy constrains the \techil{<plan>} block to a predefined set
of therapeutic strategies. This constraint is necessary for quantitative
evaluation: free-text plans cannot be aggregated into set-similarity scores.
The taxonomy evolved through four iterations to resolve a severe
distribution skew that rendered early versions unusable.

The initial taxonomy (v1) defined eight categories including
\techil{empowerment} and \techil{mastery}. Across 1{,}272 strategy picks,
these two categories consumed 87.7\% of all selections, while three
categories (\techil{social\_support}, \techil{sensory\_modulation},
\techil{gradual\_exposure}) were never chosen. The root cause was a
goal-mechanism conflation: both \techil{empowerment} and \techil{mastery}
describe a therapeutic \emph{goal} (the patient gains control) rather than
a specific \emph{mechanism} of change. Since every nightmare rescripting
intervention serves the goal of increased mastery, the model defaulted to
these labels regardless of the actual strategy employed.

Merging the two categories into a single \techil{agency} category (v2)
amplified the problem: the merged category achieved 100\% dominance. A
single super-category that applies to every rescripting scenario provides
no discriminative signal.

The resolution in v3 was to split \techil{agency} along the mechanism
dimension. \techil{Confrontation} describes external action toward the
threat (the dreamer fights, challenges, or resists). \techil{Self\_empowerment}
describes internal transformation within the dreamer (gaining abilities,
growing stronger). These two categories are orthogonal: a dreamer who
confronts a monster uses a different mechanism from one who discovers inner
strength, even though both gain mastery in the outcome.

A final revision (v4) merged \techil{emotional\_regulation} into
\techil{sensory\_modulation} after smoke testing showed that
\techil{emotional\_regulation} appeared in only 1 of 120 trials (0.4\%).
In single-turn rescripting, internal calming manifests as sensory change
(the dreamer feels calm, the environment becomes peaceful), making the
boundary between the two categories clinically indistinct at this
granularity.

The final six-category taxonomy captures concrete therapeutic mechanisms:

\begin{table}[t]
  \centering
  \caption{Strategy taxonomy (v4). Each category describes a rescripting mechanism rather than a therapeutic goal.}
  \label{tab:taxonomy}
  \begin{tabularx}{\textwidth}{lX}
    \toprule
    Category & Mechanism \\
    \midrule
    \techil{confrontation}      & The dreamer directly faces, fights, or stands up to the threat \\
    \techil{self\_empowerment}   & The dreamer gains a new ability or inner strength \\
    \techil{safety}              & A physical barrier, shelter, or protective figure shields the dreamer \\
    \techil{cognitive\_reframe}  & Existing dream elements change meaning \\
    \techil{social\_support}     & A new person, companion, or ally is introduced \\
    \techil{sensory\_modulation} & Sensory qualities shift, including internal somatic calming \\
    \bottomrule
  \end{tabularx}
\end{table}

These categories are grounded in two clinical frameworks.
\textcite{62ZMHBC3} identify six mastery dimensions in successful nightmare
rescripting (physical, social, environmental, emotional, mystical,
avoidance), all of which map to the taxonomy except avoidance.
\textcite{25UR8BHJ} provide empirical rescripting frequencies from combat
veterans, with all therapeutically productive categories covered.
Avoidance and distancing are deliberately excluded: \textcite{25UR8BHJ}
found that violence in rescripted dreams predicted worse treatment outcomes,
and \textcite{FGA5H3ZD} frame \gls{irt} as ``changing the nightmare'' rather
than escaping it. Reality markers (lucid dreaming cues) fall outside the
scope of single-turn rescripting.

The iterative development process is documented here because taxonomy design
directly shapes measured consistency. A coarse taxonomy inflates agreement
scores (every trial picks the same super-category). A fine-grained taxonomy
deflates them (minor strategy differences register as disagreements). The
sensitivity of consistency metrics to category design is revisited in the
discussion (\autoref{sec:disc-framework-design}).

% [GRAPH] Strategy frequency per taxonomy version (v1-v4)
% [FIG] 6-category taxonomy card with definitions and example phrases


\section{Evaluation Metrics}
\label{sec:meth-metrics}
% Why three methods? Each targets a distinct layer of the generation process --
%   no single metric captures the full picture of clinical stability
%   Method 1 (Plan Consistency): does the model select the same strategies?
%   Method 2 (Response Similarity): does the model produce equivalent responses?
%   Method 3 (Plan-Response Alignment): does the model do what it declared?
% [FIG] Three-level framework -- plan -> output -> alignment; inputs and outputs
%   of each method, with the clinical question each answers

Three evaluation methods target distinct layers of the generation process.
No single metric captures the full picture of clinical stability
(\autoref{sec:bg-metrics}). Method~1 asks whether the model makes consistent
therapeutic decisions. Method~2 asks whether it produces consistent
therapeutic responses. Method~3 asks whether it does what it says it will
do.

% [FIG] Three-level framework -- plan, output, alignment

% previously: Cognitive Stability (Plan Consistency) -- renamed to avoid
% theoretically loaded term (supervisor feedback 2026-03-23)
\subsection{Method 1: Plan Consistency}
\label{sec:meth-m1}
% Pairwise Jaccard similarity over strategy sets from <plan> blocks
%   J(A,B) = |A n B| / |A u B|; mean over C(20,2) = 190 trial pairs
% Strategy extraction from <plan> blocks. Validity rate as upstream quality gate
%   -- malformed plans excluded
% Why Jaccard? Set similarity is appropriate for unordered strategy
%   combinations; insensitive to labelling artefacts
% Measures: do stochastic runs produce the same therapeutic decisions?
% [Example] 10 plan sets -> pairwise Jaccard matrix -> mean score;
%   one high- and one low-consistency condition shown

Method~1 measures whether the model selects the same therapeutic strategies
across independent trials. Strategy sets are extracted from the
\techil{<plan>} block of each trial output using the fixed taxonomy from
\autoref{tab:taxonomy}. Pairwise Jaccard similarity quantifies agreement
between strategy sets:

\begin{equation}
  J(A, B) = \frac{|A \cap B|}{|A \cup B|}
\end{equation}

The mean Jaccard score is computed over all $\binom{20}{2} = 190$ trial
pairs per condition. Jaccard is appropriate because strategy combinations
are unordered sets: a model that declares \techil{confrontation /
cognitive\_reframe} has made the same clinical decision as one that declares
\techil{cognitive\_reframe / confrontation}.

Because Jaccard operates on discrete sets of one to two items drawn from
six categories, it takes only a small number of distinct values:
$\{0, 0.33, 0.5, 0.67, 1.0\}$. This discreteness means Jaccard scores are
ordinal rather than continuous, and descriptive statistics report medians
and interquartile ranges rather than means and standard deviations.

Plan validity rate serves as an upstream quality gate. A malformed
\techil{<plan>} block that cannot be parsed into valid strategy identifiers
is excluded from the Jaccard computation. The validity rate reports the
proportion of trials with parseable plans, indicating whether the model can
follow the structured output format at all.

% [Example] 10 plan sets, pairwise Jaccard matrix, mean score

% previously: Output Consistency (Semantic Stability) -- renamed for clarity
\subsection{Method 2: Response Similarity}
\label{sec:meth-m2}
% GROUNDING: builds on the metric landscape from Section 2.4
%
% Pairwise BERTScore F1 over response texts; mean over C(20,2) = 190 pairs
% Embedding model choice: DeBERTa-XLarge-MNLI (He et al., 2021)
%   Why this model?
%   - NLI fine-tuning aligns with semantic equivalence over surface matching
%   - Highest WMT16 human correlation (r = 0.778) among available options
%   - Disentangled attention handles surface-form variation in therapeutic language
%     where valid paraphrasing is the norm
% Why not BLEU/ROUGE? Surface metrics penalise valid therapeutic paraphrasing
%   (established in 2.4) -- embedding similarity is the appropriate level
% Measures: do stochastic runs produce therapeutically equivalent responses?
% [Example] Two responses from the same condition -> token alignment
%   heatmap -> F1; one high- and one low-similarity pair

Method~2 measures whether the model produces therapeutically equivalent
responses across trials, independent of strategy declarations. Pairwise
BERTScore F1 is computed over response texts (with plan blocks excluded)
across the same 190 trial pairs.

The embedding model is \modelname{DeBERTa-XLarge-MNLI} \parencite{8MQC2H2V}. This model
achieves the highest human correlation ($r = 0.778$) among more than 130
candidates on WMT16 segment-level evaluation. Its \gls{nli} fine-tuning
aligns with the semantic equivalence question: determining whether two
therapeutic responses convey the same clinical content, regardless of
surface phrasing. Its disentangled attention mechanism handles the
surface-form variation typical of therapeutic language, where a competent
response naturally varies in wording while maintaining the same clinical
intent.

Surface metrics (BLEU, ROUGE) are not used. As established in
\autoref{sec:bg-metrics}, token-overlap metrics penalise valid therapeutic
paraphrasing and cannot distinguish semantically equivalent from
semantically divergent responses when surface forms differ.

% [Example] Two responses from the same condition, alignment heatmap, F1

% previously: Plan-Output Alignment
\subsection{Method 3: Plan-Response Alignment}
% originally exploratory -- may restore label depending on results
\label{sec:meth-m3}
% GROUNDING: addresses the multi-level gap from Section 2.4
%   (no single metric captures strategic intent + semantic equivalence)
%
% LLM judge: Gemini 3.1 Pro (Google AI Studio, T=1.0), cross-family with
%   all evaluation targets. Thinking mode aids judgment accuracy.
%   T=1.0 is the Gemini 3 default; Google warns against lowering it.
%   0 = absent, 1 = partial, 2 = implemented
% Why ternary? Borrowed from clinical fidelity literature (ENACT, NIH BCC,
%   introduced in 2.1); binary scoring loses partial implementation --
%   clinically meaningful
% Why LLM judge? Only approach capable of multi-strategy alignment with
%   a reasoning trace; NLI cross-encoders ceiling at F1 ~0.55 on this task
% Rejected alternatives: NLI cross-encoders (F1 ceiling ~0.55, task mismatch,
%   no reasoning trace)
% Mitigations for known judge biases (identified in 2.4):
%   cross-model judging, CoT justification, T=1.0, transparent rubric
% Labelled exploratory: judge reliability is itself an open question
% Validation: human annotation of a random subset (~50 trials) scoring the
%   same rubric. Cohen's kappa between human and judge as reliability estimate.
%   If kappa is low, Method 3 results carry the caveat explicitly.
% Measures: does the model's response implement its declared therapeutic plan?
% [Example] Judge input + output -- plan, response, ternary verdict with
%   CoT justification; partial and absent case shown

Method~3 measures whether the model's therapeutic response implements its
declared strategy. A separate \gls{llm} judge scores each declared strategy
on a ternary scale:

\begin{itemize}
  \item \textbf{0 (absent):} The strategy is not reflected in the response.
  \item \textbf{1 (partial):} The strategy is mentioned or briefly addressed
    but not fully developed.
  \item \textbf{2 (implemented):} The strategy is clearly and substantively
    present in the response.
\end{itemize}

This ternary scale is borrowed from the clinical fidelity literature
(\autoref{sec:bg-irt}). The ENACT \parencite{GYNNP2CC} and
NIH BCC \parencite{ITT882KN} frameworks use the same three-level
gradation (absent, partial, implemented) to assess whether a human therapist
delivers an intervention correctly. Applying the same scale to \gls{llm}
outputs maintains continuity with established clinical evaluation practice.

The judge model is \modelname{Gemini 3.1 Pro}, accessed through Google AI
Studio at $T = 1.0$. \modelname{Gemini} is from a different model family
than all evaluation targets, preventing self-preference bias. Its thinking
mode supports structured reasoning about whether a response implements a
given strategy. The judge receives the strategy definitions, the declared
strategies, and the response text, and produces a score with explicit
justification for each strategy.

The judge runs at $T = 1.0$ rather than the conventional $T = 0.0$ because
Gemini~3 models are architecturally optimised for their default temperature.
Google's developer documentation warns that lowering the temperature below
1.0 can cause looping or degraded reasoning performance. Empirical evidence
supports this choice: \textcite{96K5MF6A} finds no statistically significant impact of temperature changes between 0.0 and 1.0
on \gls{llm} problem-solving accuracy. Using \modelname{Gemini 3.1 Pro} as
an \gls{llm} judge at its default temperature is consistent with recent
practice: \textcite{32QZTBPH}\footnote{Preprint, not yet peer-reviewed.} employ it as an automated
evaluator within a multi-agent assessment framework. Evaluation consistency
is instead ensured through the structured ternary rubric and explicit
scoring format.

Additional bias mitigations follow established recommendations
\parencite{LVLA9LZU}: cross-model judging ensures the judge never evaluates
its own family, \gls{cot} justification requires the judge to explain its
rating before assigning a score, and a transparent rubric anchors scores
to specific behavioural criteria.

\gls{nli} cross-encoders were considered and rejected as an alternative.
Testing showed an F1 ceiling of approximately 0.55 on the strategy-response
alignment task, insufficient for reliable scoring. Cross-encoders also
produce no reasoning trace, making it impossible to audit individual
judgements.

Method~3 is labelled exploratory. Judge reliability is itself an open
question: the judge's ternary scores are not a ground truth but a
computational proxy for clinical fidelity assessment. Validation proceeds
through human annotation of a random subset of approximately 50 trials,
scoring the same rubric. Cohen's $\kappa$ between human and judge provides
a reliability estimate. If $\kappa$ is low, Method~3 results carry the
caveat explicitly.

Smoke testing showed alignment scores near ceiling (1.0) across all trials.
If the full experiment confirms this pattern, alignment becomes a validation
check (models consistently implement what they declare) rather than a
comparative metric. This outcome would validate Method~1: if plans are
faithfully executed, then Jaccard over plan sets measures genuine
decision stability rather than label noise.

% [Example] Judge input/output -- plan, response, ternary verdict with CoT


\section{Model Selection}
\label{sec:meth-models}
% GROUNDING: regulatory framing established in 2.2
%
% EU data sovereignty as primary selection criterion -- self-hostability and EU data
%   residency are legal constraints in the reDreamAI deployment context
% All evaluation targets run in non-thinking mode for fair comparison.
%   Qwen 3.5 and Mistral Small 4 (hybrid-thinking) run with reasoning disabled.
%
% Primary subject: Mistral Large 3 (675B MoE, 41B active)
%   EU-sovereign, Apache 2.0, frontier-class
% EU-sovereign comparators: Mistral Small 4 (119B MoE), Mistral Small 3.2 (24B dense)
% Proprietary ceiling: GPT-5.4 and Sonnet 4.6 (non-thinking variants)
% Open-weight comparators across size classes and architectures
% Selection rationale: size-class diversity, dense vs MoE architecture,
%   provider diversity, EU sovereignty depth
% [FIG] Model selection table -- model, size, architecture, provider,
%   sovereignty status, role in study

\gls{eu} data sovereignty serves as the primary selection criterion for the
study's main subject. The deployment context (reDreamAI) requires a model
that can be self-hosted within \gls{eu} data residency boundaries
\parencite{2QMKF3RM}. All evaluation targets run in non-thinking mode for
fair comparison.

The \textbf{primary subject} is \modelname{Mistral Large~3}
(675B \gls{moe}, 41B active parameters, Apache~2.0). It is the \gls{eu}
sovereign flagship and the model that reDreamAI would deploy in production.
The research question centres on whether this model achieves stability
comparable to proprietary alternatives.

Two additional \gls{eu}-sovereign models complete the Mistral comparison.
\modelname{Mistral Small~4} (119B \gls{moe}, 6.5B active parameters,
Apache~2.0) is Mistral's newest release and the most efficient \gls{moe}
in the set, activating fewer parameters per token than any other model.
\modelname{Mistral Small 3.2} (24B dense, Apache~2.0) serves as a
predecessor baseline, showing whether the architectural shift from dense to
\gls{moe} affects stability.

The \textbf{proprietary ceiling} comprises two models.
\modelname{GPT-5.4} (non-thinking variant) is the strongest proprietary
model without thinking overhead. OpenAI reintroduced user-configurable
temperature with the non-reasoning variants of \modelname{GPT-5.2},
making controlled temperature comparison possible.
% TODO: source for OpenAI reintroducing temperature with GPT-5.2 non-reasoning
\modelname{Claude Sonnet~4.6} adds a second proprietary reference point;
Anthropic's character training emphasises thoughtful, safety-aware
interaction, which may influence therapeutic stability independently of raw
capability.
% TODO: add Anthropic character training citation (anthropic.com/research/claude-character)

\textbf{Open-weight comparators} span three size classes and two
architectures:

\begin{table}[t]
  \centering
  \caption{Evaluation targets. All models run in non-thinking mode via OpenRouter.}
  \label{tab:models}
  \begin{tabularx}{\textwidth}{llrX}
    \toprule
    Group & Model & Size & Selection rationale \\
    \midrule
    \multicolumn{4}{l}{\emph{EU-sovereign (Mistral, Apache~2.0)}} \\
    Primary   & \modelname{Mistral Large 3}   & 675B \gls{moe} (41B active)   & \gls{eu}-sovereign flagship. Production deployment candidate \\
    EU        & \modelname{Mistral Small 4}   & 119B \gls{moe} (6.5B active)  & Newest Mistral. Fewest active parameters of any \gls{moe} in set \\
    EU        & \modelname{Mistral Small 3.2} & 24B dense                      & Dense predecessor to Small~4 \\
    \addlinespace
    \multicolumn{4}{l}{\emph{Open-weight comparators}} \\
    \gls{moe} & \modelname{Qwen 3.5 122B}    & 122B \gls{moe} (10B active)   & Peer \gls{moe} comparator to Mistral Small~4 \\
    \gls{moe} & \modelname{Qwen 3.5 397B}    & 397B \gls{moe} (17B active)   & Large \gls{moe} comparator to Mistral Large~3 \\
    Dense     & \modelname{Qwen 3.5 27B}     & 27B dense                      & Same size class as Mistral Small 3.2 \\
    Dense     & \modelname{OLMo 3.1 32B}     & 32B dense                      & Fully open (weights, data, code). Best provenance \\
    Dense     & \modelname{Llama 3.3 70B}    & 70B dense                      & Continuity with prior efficacy study \\
    \addlinespace
    \multicolumn{4}{l}{\emph{Proprietary ceiling}} \\
    Ceiling   & \modelname{Claude Sonnet 4.6} & Proprietary                   & Character-trained for safety-aware interaction \\
    Ceiling   & \modelname{GPT-5.4}           & Proprietary                   & Strongest non-thinking proprietary model \\
    \bottomrule
  \end{tabularx}
\end{table}

The selection provides size-class diversity (24B--675B), architectural
diversity (dense \vs \gls{moe}), and provider diversity across five
vendors. \modelname{Qwen 3.5} and \modelname{Mistral Small~4} are
hybrid-thinking models that support optional reasoning. To maintain fair
comparison, reasoning is disabled for all evaluation targets.

\modelname{DeepSeek V3.2} was initially included as a large \gls{moe}
comparator but was replaced by \modelname{Qwen 3.5 397B}. DeepSeek applies
internal temperature scaling (API $T = 1.0$ maps to internal $T \approx
0.3$), which would have complicated cross-model temperature comparisons.
\modelname{Qwen 3.5 397B} occupies the same architectural niche without
this complexity. Other models considered and dropped include
\modelname{Mistral Medium 3.1} (closed weights),
\modelname{GLM-5} (a third large \gls{moe} adds little signal),
\modelname{Nemotron~3 Super} (requires $T = 1.0$ universally, complicating
temperature comparison), and \modelname{GPT-oss} (mandatory reasoning mode
cannot be disabled).


\section{Experimental Conditions}
\label{sec:meth-conditions}
% Full factorial: 6 vignettes x 5 temperatures x 20 trials x 10 models
% Temperature sweep: 0.0, 0.075, 0.15, 0.3, 0.6
%   0.0: deterministic baseline (greedy decoding)
%   0.15: Mistral vendor default
%   0.6: Qwen/Llama/OLMo vendor default, close to GPT/Sonnet 0.7
%   0.075: added post-hoc when Mistral peaked between 0.0 and 0.15
% [FIG] Condition matrix -- factorial design as a grid; total run count labelled

The experiment follows a full factorial design:

\begin{table}[t]
  \centering
  \caption{Experimental conditions. Total: $6 \times 5 \times 20 = 600$ trials per model, 6{,}000 across all ten models.}
  \label{tab:conditions}
  \begin{tabular}{ll}
    \toprule
    Dimension & Values \\
    \midrule
    Vignettes    & anxious, avoidant, cooperative, resistant, skeptic, trauma \\
    Trials       & 20 per condition \\
    Temperatures & $T \in \{0.0,\; 0.075,\; 0.15,\; 0.3,\; 0.6\}$ \\
    Models       & 10 evaluation targets (\autoref{tab:models}) \\
    \bottomrule
  \end{tabular}
\end{table}

Rather than testing only two temperature conditions, the experiment uses a
five-point temperature sweep that spans from deterministic decoding to the
vendor-recommended defaults. The scale is anchored at three points
motivated by vendor documentation:

\begin{itemize}
  \item $T = 0.0$ (greedy decoding): deterministic upper bound on stability.
    Any remaining inconsistency reflects floating-point non-determinism or
    infrastructure-level variation.
  \item $T = 0.15$: default temperature for Mistral models, taken from the
    HuggingFace generation configuration for \modelname{Mistral Small 3.2}.
    % TODO: verify Mistral Large 3 default via /models endpoint (currently
    %   verified for Small 3.2 only, see temperature_sources.md)
  \item $T = 0.6$: default temperature for \modelname{Qwen 3.5} and
    \modelname{OLMo~3.1}, also close to the standard Llama chat range
    (0.6--0.7) and the proprietary defaults (\modelname{GPT-5.4} and
    \modelname{Sonnet~4.6} default to $T = 0.7$).
    % Sources: Qwen 3.5 generation_config.json [VERIFIED],
    %   OLMo 3.1 generation_config.json [VERIFIED],
    %   Llama 3.3 -- gated HF repo, lower bound of commonly cited range
    %   [NEEDS VERIFICATION, see temperature_sources.md]
    %   GPT-5.4 -- OpenAI API docs returned 403
    %   [NEEDS VERIFICATION, see temperature_sources.md]
    %   Sonnet 4.6 -- Anthropic API docs, default 1.0, 0.7 as midpoint
    %   [VERIFIED, see temperature_sources.md]
\end{itemize}

The intermediate points $T = 0.075$ and $T = 0.3$ fill the gap between
Mistral's low default and the Qwen/Llama range. $T = 0.075$ was added
after an initial four-point run revealed that several models peak in
consistency between $T = 0.0$ and $T = 0.15$, suggesting that a small
amount of sampling variance can be beneficial (discussed in
\autoref{sec:res-plan-consistency}).

This five-point sweep replaces a conventional binary comparison
($T = 0.0$ \vs a single stochastic point). It reveals the full degradation
curve rather than a single gap, and allows each model to be evaluated at
and around its vendor-recommended operating temperature.

% [FIG] Condition matrix -- factorial design as a grid


\section{Statistical Analysis}
\label{sec:meth-stats}
% Descriptive statistics: medians/IQR for Jaccard (ordinal), means/SD for
%   BERTScore and alignment, confidence intervals across 190 trial pairs
% Spearman rank correlations for cross-method analysis (\autoref{sec:res-cross})
%   and per-model temperature effects
%   Why Spearman: rank-based, robust to non-normal distributions from
%   bounded similarity scores
% Kruskal-Wallis for model and vignette group differences
% No fixed threshold for ``stable enough.'' Results are interpreted comparatively
%   (model vs model, across temperature sweep) rather than against an absolute cutoff.
%   Clinical significance thresholds for LLM stability do not yet exist.

The analysis is primarily descriptive. For each model and temperature
condition, central tendency and dispersion are reported for all three
evaluation metrics, aggregated across vignettes.
Jaccard scores are ordinal (discrete values from a small set), so they are
summarised with medians and interquartile ranges. BERTScore F1 and alignment
scores are reported with means and standard deviations.

Spearman rank correlations serve two purposes. First, per-model
correlations between Jaccard and temperature quantify the temperature
sensitivity of each model across the five-point sweep. Second,
cross-method correlations (Method~1 \vs Method~2, Method~1 \vs Method~3,
Method~2 \vs Method~3) assess whether the three evaluation layers capture
redundant or independent information. If they diverge, the clinical
implications differ: high plan consistency with low response similarity
indicates stable strategy selections but variable execution, while the
reverse indicates different strategies producing similar surface outputs.
Spearman rather than Pearson is used because Jaccard scores are ordinal
and BERTScore distributions may be non-normal.

Kruskal-Wallis tests assess whether models and vignettes differ
significantly on each metric. These are exploratory rather than
confirmatory: the study is designed to describe stability patterns, not to
test a preregistered hypothesis.

No fixed threshold defines ``stable enough.'' Clinical significance
thresholds for \gls{llm} stability do not yet exist in the literature.
Results are interpreted comparatively: model against model, across the
temperature sweep. The random Jaccard baseline (expected value for random
2-of-6 selections, approximately 0.2) provides a lower reference point.
