\chapter{Implementation}
\label{chap:implementation}

% Role: translates Chapter 3 design decisions into concrete technical choices.
% Each decision here is a consequence of a methods requirement.

This chapter describes how the methods from \autoref{chap:methods} are
realised in software. The pipeline is implemented in Python and structured
around three concerns: a provider-agnostic \gls{llm} interface, a
config-driven experiment runner, and a post-hoc aggregation pipeline. All
source code is available in the project repository.


\section{Software Architecture}
\label{sec:impl-architecture}
% Python async-first pipeline; YAML config; Pydantic validation
%   Why async? 10 parallel trials per condition -- sequential would be prohibitive
%   Why YAML? Every condition is a config parameter -- reproducibility without
%     code changes; full experiment defined declaratively
%   Why Pydantic? Malformed <plan> blocks must fail at parse time --
%     silent propagation would corrupt downstream metrics
% Multi-provider LLM abstraction via OpenRouter and Google AI Studio
%   OpenRouter for all evaluation targets, Google AI Studio for Gemini judge
%   Enables model swap without changing evaluation logic -- critical for
%   cross-model comparison
% [FIG] Component diagram -- modules, interfaces, provider abstraction layer

The pipeline follows an async-first design. Every \gls{llm} call is
non-blocking, and trials within a condition run concurrently via
\techil{asyncio.gather()}. This is a direct consequence of the factorial
design: 360 trials per model would be prohibitive if executed sequentially.
Parallel execution reduces wall-clock time from hours to minutes per model.

\paragraph{Configuration.}
Two YAML files define the full experiment declaratively.
\techil{models.yaml} specifies provider endpoints, model identifiers, role
assignments, and evaluation targets. \techil{experiment.yaml} controls
sampling parameters (trial count, temperatures, maximum token length),
metric selection, and output paths. Changing the evaluation target or
temperature requires editing a single YAML field rather than modifying
code. This separation ensures that experiment conditions are reproducible:
the configuration snapshot saved with each run fully specifies how to
recreate it.

\paragraph{Data validation.}
Pydantic models enforce type safety at every data boundary. The
\techil{Message} and \techil{Conversation} classes validate all fields on
construction and serialisation. When a frozen history is loaded from disk,
\techil{Conversation.from\_dict()} uses \techil{model\_validate()} with
\techil{extra="ignore"}, silently discarding metadata fields that do not
belong to the conversation schema. This is intentional: frozen histories
may carry generation-time metadata (timestamps, model versions) that is
irrelevant to evaluation.

The same validation applies to plan extraction. The
\techil{extract\_plan\_strategies()} function parses the \techil{<plan>}
block from each trial output and validates each strategy identifier against
the taxonomy. A malformed plan that cannot be parsed into valid strategy
identifiers fails at extraction time rather than propagating downstream
where it would silently corrupt Jaccard scores.

\paragraph{Provider abstraction.}
% Multi-provider LLM abstraction via OpenRouter and Google AI Studio
%   OpenRouter for all evaluation targets, Google AI Studio for Gemini judge
%   Enables model swap without changing evaluation logic -- critical for
%   cross-model comparison
A single \techil{LLMProvider} class wraps an \techil{AsyncOpenAI} client
and exposes one async method: \techil{generate(messages, **kwargs)},
returning a tuple of response content and token usage. All six evaluation
targets, the patient agent, and the router use OpenRouter through this
interface. The \modelname{Gemini} judge uses Google AI Studio through the
same interface, since Google AI Studio exposes an OpenAI-compatible
endpoint. Adding a new model or provider requires a configuration entry,
not a code change.

A factory function (\techil{create\_provider()}) resolves role names
(\eg ``therapist'', ``judge'') to concrete provider configurations by
looking up the active model assignment in \techil{models.yaml}. For
experiment-tier runs, the judge automatically upgrades from
\modelname{Gemini Flash Lite} (used during development) to
\modelname{Gemini 3.1 Pro}.

% [FIG] Component diagram -- modules, interfaces, provider abstraction layer

The codebase is organised into five packages:

\begin{table}[t]
  \centering
  \caption{Package structure. Each package maps to a distinct concern in the pipeline.}
  \label{tab:packages}
  \begin{tabularx}{\textwidth}{lX}
    \toprule
    Package & Responsibility \\
    \midrule
    \techil{src/agents/}     & Agent implementations (patient, therapist, router). Each agent extends a \techil{BaseAgent} class that handles message formatting, system prompt injection, and provider delegation \\
    \techil{src/core/}       & Data models (\techil{Message}, \techil{Conversation}), stage and language enums, YAML and prompt loaders \\
    \techil{src/llm/}        & Provider abstraction. Single-class interface to all \gls{llm} backends \\
    \techil{src/stacks/}     & Orchestration. \techil{GenerationStack} runs the three-agent dialogue loop. \techil{EvaluationStack} runs fused plan-and-response generation from frozen histories \\
    \techil{src/evaluation/} & Experiment runner, multi-trial sampler, and all three evaluation metrics (Jaccard, BERTScore, \gls{llm} judge alignment) \\
    \bottomrule
  \end{tabularx}
\end{table}


\section{Experiment Execution}
\label{sec:impl-execution}
% ExperimentRun orchestration with structured output
% Config -> frozen history -> parallel trials -> output bundle
% Multi-trial sampling with parallel execution
% Artefacts per run: config.yaml, frozen_history.json, trials/,
%   metrics.json, judgments.json
%   Why this structure? Frozen history is the only irreplaceable artefact;
%   metrics and judgments are recomputable if evaluation logic changes
% [Example] Output directory tree for one complete experiment run

An \techil{ExperimentRun} object manages a single evaluation run: one
model, one vignette, one slice, one temperature, ten trials. On
initialisation it creates a timestamped output directory and saves the
configuration snapshot and frozen history.

Execution proceeds in three phases:

\begin{enumerate}
  \item \textbf{Sampling.} A \techil{Sampler} wraps the
    \techil{EvaluationStack} and runs ten independent trials. Each trial
    calls the therapist model once (fused mode), producing a
    \techil{<plan>} block and a therapeutic response. Trials run in
    parallel via \techil{asyncio.gather()} with built-in retry logic for
    rate-limit errors (HTTP~429) using exponential backoff.
  \item \textbf{Metric computation.} After all trials complete, the run
    computes all three evaluation metrics. Method~1 (Jaccard) and Method~2
    (BERTScore) are computed locally. Method~3 (alignment) issues parallel
    async calls to the \modelname{Gemini} judge, with a 120-second timeout
    per trial to handle unresponsive judge calls gracefully.
  \item \textbf{Persistence.} Results are saved to the run directory.
\end{enumerate}

Each run produces a self-contained output bundle:

% TODO: consider replacing with dirtree package or a proper figure
\begin{verbatim}
experiments/runs/{timestamp}_{model}_{vignette}/
    config.yaml              # run parameters (model, temp, slice)
    frozen_history.json      # input conversation (irreplaceable)
    trials/
        trial_01.json        # plan + response + token usage
        ...
        trial_10.json
    metrics.json             # Jaccard, BERTScore, validity, alignment
    judgments.json            # raw judge outputs with reasoning
\end{verbatim}

The frozen history is the only irreplaceable artefact in this structure.
If the evaluation logic changes (a bug fix in Jaccard computation, a new
judge prompt), metrics and judgements can be recomputed from the saved
trials. Trials themselves can be recomputed from the frozen history plus
the configuration, though at the cost of additional \gls{api} calls.

The plan extraction logic handles several output formats that different
models produce. The primary pattern matches a closed \techil{<plan>}
tag. Fallback patterns handle unclosed tags, bare tag-style strategy
names, and plain-text first-line declarations. This robustness is
necessary because the six evaluation targets do not all follow the
structured output format identically, despite receiving the same system
prompt.

% [Example] Output directory tree for one complete experiment run


\section{Aggregation Pipeline}
\label{sec:impl-aggregation}
% Cross-experiment result collection
% Collects all model x vignette x slice x temperature results into one
%   analysis frame
% Produces per-model, per-vignette, per-slice, per-temperature breakdowns
%   Cross-method correlation (5.5) requires all three metric types aligned
%   on the same condition index -- this pipeline produces that alignment

After all experiment runs complete, an aggregation script collects results
into a single analysis frame. The script iterates over all run directories,
loads each \techil{config.yaml} and \techil{metrics.json}, and produces one
row per run indexed by model, vignette, slice, and temperature.

Strategy counts are expanded into six separate columns (one per taxonomy
category), enabling per-strategy frequency analysis. Alignment scores per
strategy are similarly expanded. The resulting DataFrame is saved as a CSV
file (\techil{stats/data/all\_runs.csv}) that serves as the input for all
downstream analysis and visualisation scripts.

This unified frame is a prerequisite for the cross-method analysis in
\autoref{sec:res-cross}. Spearman correlations between Method~1, Method~2,
and Method~3 require all three metrics aligned on the same condition index.
Without a single frame that maps each condition to its Jaccard, BERTScore,
and alignment scores, cross-method comparison would require manual alignment
across separate output files.

Downstream scripts in \techil{stats/scripts/} consume this CSV to produce
descriptive statistics (\techil{descriptives.py}), statistical tests
(\techil{tests.py}), and all five result figures. The separation between
aggregation and analysis allows each step to be rerun independently: if a
figure script is modified, only visualisation reruns, not the full
aggregation.
