\chapter{Discussion}
\label{chap:discussion}

\section{Interpreting the Three-Level Framework}
\label{sec:disc-framework-interpretation}
% What the three methods together reveal that no single method could show alone
% Cases where the levels align vs. diverge -- and what each divergence type means
% The framework as a diagnostic tool: which layer is the source of instability?
% [FIG] Stability failure taxonomy -- which divergence pattern indicates which
%   type of clinical risk

The three evaluation methods were designed to capture distinct aspects of
therapeutic stability (\autoref{sec:meth-metrics}). The results confirm
that they do: the cross-method correlations are moderate at best
($\rho = 0.560$ for Jaccard--BERTScore, $\rho = 0.285$ for
Jaccard--Alignment). Had any two methods been strongly correlated, one
would be redundant. Instead, each method reveals information that the
others miss.

The most informative finding is the dissociation between plan consistency
and alignment. Temperature increases reduce Jaccard (models select
different strategies) but do not reduce alignment (models implement
whichever strategy they select equally well). This pattern localises the
source of instability to the \emph{decision-making} layer rather than the
\emph{execution} layer. A model at high temperature does not produce
incoherent or unfaithful responses. It produces coherent responses to a
different therapeutic plan.

This distinction has practical value as a diagnostic tool. If a deployment
shows unstable therapeutic behaviour, the framework can identify where the
instability originates. High Jaccard with low alignment would indicate
that the model selects consistent strategies but fails to implement them
(an execution problem). Low Jaccard with high alignment, which is the
dominant pattern observed here, indicates that the model implements its
plans faithfully but selects different plans across runs (a decision
problem). Each pattern implies a different remediation strategy:
execution problems suggest prompt engineering or fine-tuning, while
decision problems suggest temperature tuning or strategy constraints.

BERTScore occupies a more limited role than initially anticipated. It
correlates moderately with Jaccard but cannot distinguish between
clinically different plans when the surface text remains similar. The
slice depth analysis confirms this: Jaccard varies across conversation
depths while BERTScore stays flat (\autoref{fig:slice-depth}). A model
that switches from \techil{confrontation} to \techil{cognitive\_reframe}
changes its therapeutic approach, but the response text remains similar
enough that BERTScore does not detect the difference. BERTScore therefore
serves primarily as a sanity check for response coherence rather than as
an independent stability measure.


\section{Cross-Model Comparison}
\label{sec:disc-models}
% Mistral Large 3 vs. GPT-5.4 and open-weight peers across all three methods
% Sovereignty dimension: does the EU flagship match proprietary models?
% Does self-hostability come at a stability cost? -- the sovereignty tradeoff
% Size-class effects: small (24-32B) vs. mid (70B) vs. large MoE (~40B active)
%   vs. proprietary frontier. Does more parameters mean more stability?
% [GRAPH] Summary: all models x all three metrics x both temperatures --
%   the central comparative result

The central question of this thesis is whether the \gls{eu}-sovereign
flagship, \modelname{Mistral Large~3}, achieves therapeutic stability
comparable to proprietary alternatives (\autoref{fig:jaccard}).

At the vendor-recommended temperature ($T = 0.15$),
\modelname{Mistral~Large~3} achieves $J = 1.000$, matching
\modelname{GPT-5.4} and \modelname{Sonnet~4.6}. At its peak ($T = 0.075$),
it also reaches $J = 1.000$. The sovereignty tradeoff, at low
temperatures, is negligible: the \gls{eu} flagship matches the proprietary
ceiling on plan consistency.

The tradeoff becomes visible at higher temperatures. At $T = 0.6$,
\modelname{Mistral~Large~3} drops to $J = 0.652$, while the proprietary
models maintain $J = 1.000$. However, $T = 0.6$ is well above Mistral's
recommended operating range. A deployment that follows vendor guidance
($T \leq 0.15$) would not encounter this degradation.

\modelname{Mistral~Small~3.2} (24B dense) is the most stable open-weight
model in the set, maintaining $J = 1.000$ through $T = 0.3$ and showing no
statistically significant temperature sensitivity.

\modelname{Mistral~Small~4} presents the most pronounced version of the
greedy-suboptimality pattern. At $T = 0.0$ it scores only $J = 0.722$,
then jumps to $J = 1.000$ at $T = 0.075$. A small amount of sampling
variance helps the model settle on a consistent strategy rather than
oscillating between candidates under greedy decoding.
\modelname{Mistral~Large~3} ($0.933 \to 1.000$) and
\modelname{Llama~3.3~70B} ($0.756 \to 0.881$) show the same direction.

Among the open-weight comparators, parameter count does not straightforwardly
predict stability. \modelname{Qwen~3.5~122B} (10B active) outperforms
\modelname{Qwen~3.5~397B} (17B active) at every temperature point.
\modelname{OLMo~3.1~32B} matches larger models at low temperatures but
degrades significantly at higher ones ($\rho = -0.632$, strongest
temperature sensitivity in the set). The Qwen~3.5 family shows a
consistent pattern across its three size classes: the mid-size 122B model
is the most stable, while the smallest (27B) and largest (397B) are less
consistent.

The proprietary models (\modelname{GPT-5.4}, \modelname{Sonnet~4.6})
form a distinct tier. Their perfect Jaccard across all temperatures
suggests that character training, safety-layer alignment, or other
proprietary techniques create a stability floor that sampling temperature
cannot disrupt. Whether this reflects genuine robustness or a narrower
output distribution (fewer strategies available to vary) cannot be
determined from these results alone.


\section{Temperature and Clinical Deployment}
\label{sec:disc-temperature}
% Greedy decoding as an upper-bound baseline -- theoretical best case
% Stochastic drift patterns: which models degrade most across temperature sweep?
% T=0.075 sweet spot: why greedy decoding is not always optimal
% Clinical significance of measured drift patterns
% Practical deployment recommendations: vendor defaults vs. clinical stability
% [GRAPH] Temperature degradation curves per model -- the stochastic cost

The five-point temperature sweep reveals that the relationship between
temperature and stability is not monotonic for all models. Three models
achieve strictly higher Jaccard at $T = 0.075$ than at $T = 0.0$, meaning
greedy decoding is actively counterproductive for them.

Three models achieve their peak Jaccard at $T = 0.075$ rather than
$T = 0.0$: \modelname{Mistral~Small~4} ($0.722 \to 1.000$),
\modelname{Mistral~Large~3} ($0.933 \to 1.000$), and
\modelname{Llama~3.3~70B} ($0.756 \to 0.881$). This finding has a
practical implication: clinicians or engineers deploying these models
should not default to $T = 0.0$ for maximum stability. A small amount of
sampling variance appears to help the model resolve between competing
strategy candidates, producing more consistent selections than pure
greedy decoding.

The pooled Spearman correlation ($\rho = -0.228$, $p < 0.001$,
\autoref{tab:temp-spearman}) confirms that higher temperature generally
reduces plan consistency, but the effect is weak. Per-model analysis
reveals that only three models show statistically significant degradation: \modelname{OLMo~3.1~32B}, \modelname{Qwen~3.5~27B},
and \modelname{Qwen~3.5~122B}. The remaining seven models, including the
entire Mistral family, show no significant temperature effect on Jaccard.

For clinical deployment, these results suggest a practical recommendation:
operate at or near the vendor-recommended temperature. Models designed for
a specific temperature range tend to be most stable within it. Pushing a
model to $T = 0.0$ in the hope of maximising determinism can be
counterproductive.

Alignment is temperature-independent across all models ($\rho = -0.053$,
$p = 0.361$). This means that the clinical cost of higher temperature is
limited to strategy \emph{selection} variability. The model does not become
worse at \emph{executing} its chosen strategy. A patient receiving care
from a high-temperature model would encounter different therapeutic
approaches across sessions, but each individual session would be internally
coherent.


\section{Vignette-Dependent Behaviour}
\label{sec:disc-vignettes}
% Which patient profiles challenge model consistency most: resistant, trauma
% Strategy selection patterns across clinical presentations -- are some
%   strategies inherently harder to maintain consistently?
% Implications for deployment: not all patient profiles are equally safe
% Implications for patient-specific adaptation
% [GRAPH] Per-vignette stability ranking across all three methods

The significant vignette effect on Jaccard ($p = 0.007$) and BERTScore
($p = 0.015$) indicates that some patient profiles are harder to treat
consistently than others. This is clinically relevant: if a model is stable
for cooperative patients but variable for resistant or trauma-presenting
patients, deployment safety depends on the patient population being served.

This finding suggests that stability evaluation should not rely solely on
aggregate metrics. Per-vignette reporting, as provided in this study's
appendix, is necessary to identify the conditions under which a model's
consistency breaks down. A model that appears stable on average may be
unreliable for the specific patient profiles that need the most careful
treatment.

Full per-vignette analysis is provided in Appendix~D.

% TODO: reference Fig D.3 (vignette heatmaps) when appendix written


\section{Evaluation Framework Reflections}
\label{sec:disc-framework-design}
% Limitation: no reproducibility seed was set. Nine of ten models accept a
%   seed parameter (Sonnet does not), but all vendors describe it as best
%   effort. Trial repetition captures variance directly (Yuan et al., 2025;
%   Cohn & Blackwell, 2025).
% Strengths: hierarchical decomposition, reproducibility, clinical grounding
% What the taxonomy design choices reveal -- and what they constrain
%   Taxonomy sensitivity: how category design shapes measured consistency
% What declared intent can and cannot claim about model cognition
%   Plan-as-declared-intent: what this framing enables and what it cannot claim
% Generalisability: can this framework apply to CBT, DBT, other manualized protocols?
% Broader perspective: framework as reusable validation template for
%   protocol-driven AI agents. Computational evaluation complements but
%   does not substitute clinical validation.

\paragraph{Strengths.}
The framework's hierarchical structure proved useful in practice. By
measuring stability at three layers, the analysis could identify that
instability is concentrated at the decision-making level (variable
strategy selection) rather than the execution level (unfaithful
implementation). A single-metric evaluation using only BERTScore, for
instance, would have missed the plan-level instability entirely, as
BERTScore was insensitive to strategy changes
(\autoref{sec:res-similarity}).

The frozen history design ensures strict reproducibility: every trial
receives identical input, and any observed variance is attributable to the
model under test. The 20-trial repetition provides statistical coverage of
the output distribution rather than relying on a single evaluation
\parencite{WA4QD3UE}.

\paragraph{Taxonomy sensitivity.}
The measured consistency scores are sensitive to taxonomy design. A coarser
taxonomy (fewer, broader categories) inflates Jaccard scores because more
strategy selections fall into the same category. A finer taxonomy deflates
them. The six-category taxonomy used here (\autoref{tab:taxonomy})
represents a balance between clinical granularity and discriminative power,
developed through four iterations that resolved a severe distribution skew
(\autoref{sec:meth-taxonomy}). Different taxonomy choices would produce
different absolute Jaccard scores. The \emph{relative} ordering of models
is more robust to taxonomy changes than the absolute values, because all
models are evaluated against the same category set.

\paragraph{Declared intent, not cognition.}
The \techil{<plan>} block is framed as declared intent rather than a
window into model cognition (\autoref{sec:meth-plan}). This framing has
a limitation: it cannot distinguish between a model that genuinely
decides on a strategy and then implements it, and a model that generates
a plausible-sounding plan after having already committed to a response
direction. The high alignment scores observed across most models are
consistent with either interpretation. What the plan \emph{can} claim is
that strategy declarations vary across runs, and this variation predicts
variable therapeutic behaviour regardless of the underlying mechanism.

\paragraph{Generalisability.}
The three-level framework is not specific to \gls{irt}. Any manualized
therapeutic protocol with identifiable strategy categories could be
evaluated using the same structure: extract declared strategies (Method~1),
compare response texts (Method~2), and verify plan-response alignment
(Method~3). Candidate protocols include \gls{cbt} (with its established
technique taxonomy), dialectical behaviour therapy, and motivational
interviewing. The framework would require a new strategy taxonomy for each
protocol, but the evaluation infrastructure and metrics transfer directly.

\paragraph{Ethical scope.}
This framework measures whether a model produces \emph{consistent}
therapeutic responses, not whether those responses are \emph{clinically
appropriate}. A model that consistently selects the wrong strategy scores
high on plan consistency. Stability is a necessary precondition for
clinical deployment, as stated in \autoref{sec:intro-ai-therapy}, but it
is not sufficient. All evaluations use synthetic patients and no human
participants are involved. Clinical validation through human expert
evaluation remains necessary before any deployment to real patients.

\paragraph{Methodological constraints.}
No reproducibility seed was set, as not all vendors support one. Even
where seeds are available, they provide best-effort determinism at most
\parencite{HITR7PL4}. The trial repetition design captures stochastic
variance directly.

The experiment evaluates at a single conversation depth
(\techil{slice\_2}), chosen as a conservative estimate based on the
preliminary depth analysis (\autoref{sec:meth-frozen}). Results at other
depths may differ. All models were accessed through OpenRouter, so
provider-side batching and load balancing may contribute to the
non-determinism observed at $T = 0.0$, though this mirrors realistic
deployment conditions.

Further limitations regarding scope, vignette coverage, judge validation,
and taxonomy constraints are discussed in \autoref{sec:concl-limitations}.
