\chapter{Conclusion}
\label{chap:conclusion}

\section{Summary of Findings}
\label{sec:concl-summary}
% Key results across all three evaluation levels
% Direct answer to the research question: how consistent are sovereign LLMs
%   in a structured therapeutic protocol -- and what does that mean clinically?

This thesis asked how consistent \gls{llm} therapeutic responses are
across independent runs within a structured \gls{irt} protocol. Three
evaluation methods were applied to ten models across 6{,}000 trials
spanning five temperature conditions.

The main findings are:

\begin{enumerate}
  \item \textbf{The \gls{eu}-sovereign flagship matches proprietary models
    at recommended temperatures.} \modelname{Mistral Large~3} achieves
    $J = 1.000$ at $T = 0.075$ and $T = 0.15$, equal to \modelname{GPT-5.4}
    and \modelname{Sonnet~4.6}. The sovereignty tradeoff in therapeutic
    stability is negligible when the model operates within its vendor-recommended
    range.

  \item \textbf{Instability is a decision-making problem, not an execution
    problem.} Temperature reduces plan consistency (Jaccard) but does not
    reduce plan-response alignment. Models select different strategies at
    higher temperatures but implement whichever strategy they select equally
    well.

  \item \textbf{Greedy decoding is not always optimal.} Three models achieve
    higher consistency at $T = 0.075$ than at $T = 0.0$. A small amount of
    sampling variance can help a model settle on a consistent strategy.

  \item \textbf{The three evaluation levels capture distinct information.}
    Cross-method correlations are moderate at best ($\rho = 0.285$ to
    $0.560$). BERTScore is insensitive to plan-level changes. No single
    metric captures the full picture of therapeutic stability.

  \item \textbf{Patient profile matters.} A significant vignette effect
    ($p = 0.007$) means that stability evaluations based on aggregate
    metrics alone may miss per-profile weaknesses.
\end{enumerate}

Regarding the two contributions stated in \autoref{sec:intro-objectives}:
the three-level framework (Contribution~1) fulfilled its design goal by
identifying that instability is concentrated at the strategy selection
layer, a finding that neither BERTScore nor alignment alone could have
revealed. The cross-model comparison (Contribution~2) demonstrates that
the \gls{eu}-sovereign \modelname{Mistral Large~3} achieves stability
comparable to proprietary models at its recommended operating temperature,
supporting its viability for clinical deployment under sovereignty
constraints.


\section{Limitations}
\label{sec:concl-limitations}
% Scope: single protocol, single phase (IRT rescripting only)
% Simulation: synthetic patients differ from real clinical interaction in
%   ways that cannot be fully controlled
% Six vignettes sample the difficulty range but cannot claim exhaustive
%   coverage of patient variability
% Method 3 is exploratory -- judge reliability is unverified and should
%   not be treated as ground truth
% Fixed taxonomy constrains the measurable strategy space -- unmeasured
%   strategies may show different stability patterns
% Computational stability does not equal clinical safety

\paragraph{Scope.}
The study evaluates a single phase (rescripting) of a single protocol
(\gls{irt}). Stability in other \gls{irt} stages or other therapeutic
protocols may differ. The rescripting phase was selected because it
requires active strategy decisions, but this focus limits the
generalisability of the absolute scores reported.

\paragraph{Synthetic patients.}
All patient interactions are simulated by an \gls{llm}. Synthetic patients
follow scripted behavioural profiles and do not exhibit the unpredictable
responses of real patients. The vignettes are designed to sample the
difficulty range (\autoref{sec:meth-dialogue}), but six profiles cannot
claim exhaustive coverage of clinical variability. A cooperative synthetic
patient may underestimate the challenges a model faces with a real patient
whose behaviour deviates from any template.

\paragraph{Ethical considerations.}
This study evaluates \gls{llm} stability using synthetic patients, not
real clinical interactions. No human participants were involved. The
results measure computational consistency and do not constitute evidence
of clinical safety or efficacy. \gls{irt} targets nightmares rather than
psychiatric diagnoses, which limits clinical risk relative to full
psychotherapy, but any deployment to real patients would require clinical
validation beyond what this framework provides.

\paragraph{Judge validation.}
The \gls{llm} judge (\modelname{Gemini 3.1 Pro}) has not been validated
against human annotations. Method~3 results are a computational proxy for
clinical fidelity assessment and should be interpreted accordingly. The
planned Cohen's $\kappa$ validation remains outstanding.

\paragraph{Taxonomy constraints.}
The six-category strategy taxonomy constrains what can be measured. If
models employ therapeutic strategies that fall outside these categories,
that variation is invisible to Method~1. The taxonomy was developed through
iterative refinement (\autoref{sec:meth-taxonomy}), but different category
boundaries would produce different absolute consistency scores.

\paragraph{Consistency is not quality.}
A model that consistently selects an inappropriate strategy scores high on
plan consistency. Stability is a necessary but not sufficient condition for
clinical deployment.


\section{Future Work}
\label{sec:concl-future}
% Full IRT protocol evaluation -- beyond the rescripting phase
% Extension to other manualized therapies (CBT, DBT, PE)
% Longitudinal drift: does stability degrade across model version updates?
% Multi-language evaluation -- German therapeutic context for reDreamAI
% Clinical validation: human therapist ratings as external ground truth
% Replication with future model generations to track whether stability
%   improves across release cycles

Several directions follow from this work.

\paragraph{Full protocol evaluation.}
Extending the evaluation beyond the rescripting phase to all five \gls{irt}
stages would test whether the stability patterns observed here hold when
the therapist model faces different types of clinical decisions (listening,
summarising, rehearsing).

\paragraph{Other manualized protocols.}
The three-level framework transfers to any protocol with identifiable
strategy categories. \gls{cbt}, dialectical behaviour therapy, and
motivational interviewing are natural candidates. Each would require a
domain-specific taxonomy but could reuse the evaluation infrastructure.

\paragraph{Longitudinal drift.}
Model vendors update weights and serving infrastructure without notice.
Repeating the evaluation across model versions would reveal whether
stability improves, degrades, or fluctuates across release cycles.

\paragraph{Multi-language evaluation.}
reDreamAI targets German-speaking users. Evaluating therapeutic stability
in German would test whether the patterns observed in English transfer
across languages, or whether language-specific training data introduces
different stability profiles.

\paragraph{Clinical validation.}
Human therapist ratings on the same ternary rubric would validate Method~3
and provide an external ground truth for the computational framework.
Comparing human and \gls{llm} judge scores would quantify the reliability
of automated fidelity assessment.
