% Measure-AI-Drift Thesis
% Bachelor's Thesis: LLM Stability in Therapeutic Contexts
%
% Compile with: latexmk -pdf main.tex
% Or: pdflatex main.tex && biber main && pdflatex main.tex && pdflatex main.tex

\documentclass[11pt,
  paper=a4,
  bibliography=totocnumbered,
  captions=tableheading,
  BCOR=10mm
]{scrreprt}

% --- Encoding and fonts ---
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% --- Language ---
\usepackage[english]{babel}

% --- Spacing ---
\usepackage[onehalfspacing]{setspace}

% --- Quoting ---
\usepackage{csquotes}

% --- Math ---
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{mathtools}

% --- Floats and figures ---
\usepackage[section]{placeins}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{floatrow}
\floatsetup[table]{style=plaintop}
\graphicspath{{figures/}{../visualizations/}}

% --- Tables ---
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{multirow}

% --- Typography ---
\usepackage[all]{nowidow}

% --- Hyperlinks ---
\usepackage[
  colorlinks=true,
  linkcolor=black,
  citecolor=blue!60!black,
  urlcolor=blue!70!black
]{hyperref}

% --- Code listings ---
\usepackage{xcolor}
\usepackage{listings}
\lstset{
  basicstyle=\ttfamily,
  keywordstyle=\color{black}\bfseries\underbar,
  identifierstyle=,
  commentstyle=\color{white},
  stringstyle=\ttfamily,
  showstringspaces=false,
  numbers=left,
  numberstyle=\tiny,
  stepnumber=1
}

% JSON syntax highlighting
\colorlet{json:punct}{red!60!black}
\colorlet{json:numb}{magenta!60!black}
\definecolor{json:delim}{RGB}{20,105,176}
\lstdefinelanguage{json}{
  basicstyle=\linespread{0.9}\ttfamily,
  showstringspaces=false,
  breaklines=true,
  breakindent=4pt,
  breakautoindent=false,
  literate=
   *{0}{{{\color{json:numb}0}}}{1}
    {1}{{{\color{json:numb}1}}}{1}
    {2}{{{\color{json:numb}2}}}{1}
    {3}{{{\color{json:numb}3}}}{1}
    {4}{{{\color{json:numb}4}}}{1}
    {5}{{{\color{json:numb}5}}}{1}
    {6}{{{\color{json:numb}6}}}{1}
    {7}{{{\color{json:numb}7}}}{1}
    {8}{{{\color{json:numb}8}}}{1}
    {9}{{{\color{json:numb}9}}}{1}
    {:}{{{\color{json:punct}{:}}}}{1}
    {,}{{{\color{json:punct}{,}}}}{1}
    {\{}{{{\color{json:delim}{\{}}}}{1}
    {\}}{{{\color{json:delim}{\}}}}}{1}
    {[}{{{\color{json:delim}{[}}}}{1}
    {]}{{{\color{json:delim}{]}}}}{1},
}

% Python syntax highlighting
\definecolor{py:keyword}{RGB}{0,128,0}
\definecolor{py:string}{RGB}{186,33,33}
\definecolor{py:comment}{RGB}{128,128,128}
\lstdefinestyle{python}{
  language=Python,
  basicstyle=\linespread{0.9}\ttfamily\small,
  keywordstyle=\color{py:keyword}\bfseries,
  stringstyle=\color{py:string},
  commentstyle=\color{py:comment}\itshape,
  showstringspaces=false,
  breaklines=true,
  numbers=left,
  numberstyle=\tiny\color{gray},
}

% --- Colored boxes (for prompt examples, model outputs) ---
\usepackage[skins,breakable]{tcolorbox}
\tcbset{boxrule=0.5pt}

% --- Bibliography (APA style for psychology-adjacent thesis) ---
\usepackage[
  backend=biber,
  style=apa,
  sorting=nyt
]{biblatex}
\addbibresource{references.bib}

% --- Glossary and acronyms ---
\usepackage[automake, nopostdot, nonumberlist]{glossaries}
\makeglossaries

% --- Misc ---
\usepackage{authoraftertitle}

%%% Custom commands %%%

% Model names in small caps
\newcommand{\modelname}[1]{\textsc{#1}}

% Inline code
\newcommand{\codeil}[1]{\lstinline{#1}}{}
\newcommand{\techil}[1]{\texttt{#1}}

% Shorthands
\newcommand{\ie}{i.\,e.~}
\newcommand{\eg}{e.\,g.~}
\newcommand{\vs}{vs.~}

% Reference helpers
\newcommand{\figref}[1]{(cf.~Figure \ref{#1})}
\newcommand{\figureref}[1]{Figure \ref{#1}}
\newcommand{\tabref}[1]{(cf.~Table \ref{#1})}
\newcommand{\tableref}[1]{Table \ref{#1}}

% Figure width
\def \figwidth {0.9\linewidth}

% Capitalized autoref names
\renewcommand*{\chapterautorefname}{Chapter}
\renewcommand*{\sectionautorefname}{Section}

% Signature line for declaration
\newcommand{\namesigdate}[1][5cm]{%
  \vspace{5cm}
  {\setlength{\parindent}{0cm}
  \begin{minipage}{0.3\textwidth}
    \hrule
    \vspace{0.5cm}
    {\small city, date}
  \end{minipage}
  \hfill
  \begin{minipage}{0.3\textwidth}
    \hrule
    \vspace{0.5cm}
    {\small signature}
  \end{minipage}
  }
}

%%% Acronym definitions %%%
\newacronym{llm}{LLM}{Large Language Model}
\newacronym{irt}{IRT}{Imagery Rehearsal Therapy}
\newacronym{moe}{MoE}{Mixture of Experts}
\newacronym{eu}{EU}{European Union}
\newacronym{api}{API}{Application Programming Interface}
\newacronym{json}{JSON}{JavaScript Object Notation}
\newacronym{cbt}{CBT}{Cognitive Behavioral Therapy}
\newacronym{rlhf}{RLHF}{Reinforcement Learning with Human Feedback}
\newacronym{gdpr}{GDPR}{General Data Protection Regulation}
\newacronym{ptsd}{PTSD}{Post-Traumatic Stress Disorder}
\newacronym{cot}{CoT}{Chain-of-Thought}
\newacronym{nli}{NLI}{Natural Language Inference}
\newacronym{bdi}{BDI}{Brief Dreamer Interview}
\newacronym{nlp}{NLP}{Natural Language Processing}

%%% Document metadata %%%
\title{Measuring Drift in Therapeutic AI: A Stability-Based Evaluation of Sovereign LLMs in Nightmare Therapy}
\author{Daniel Menzel}
\date{\today}

\begin{document}

% --- Title page ---
\begin{titlepage}
  \begin{flushleft}
    Universit\"at Osnabr\"uck\\
    Fachbereich Humanwissenschaften\\
    Institute of Cognitive Science
  \end{flushleft}

  \vspace{2cm}
  \centering{
    Bachelor's Thesis\vspace{1cm}\\
    \textbf{\Large{\MyTitle}}
    \vspace{1cm}\\
    \begin{tabular}{c}
      \MyAuthor                             \\
      979 379                               \\
      Bachelor's Program Cognitive Science  \\
      Winter 2025/2026
    \end{tabular}}
  \vspace{1cm}

  \begin{tabular}{ll}
    First supervisor:  & Moritz Hartstang                \\
                       & Institute of Cognitive Science  \\
                       & Osnabr\"uck                     \\\\
    Second supervisor: & Prof.\ Dr.\ Sebastian Musslick  \\
                       & Institute of Cognitive Science  \\
                       & Osnabr\"uck
  \end{tabular}

\end{titlepage}


% --- Declaration of Authorship ---
\chapter*{Declaration of Authorship}
The content of this thesis represents my own knowledge, my own understanding and my own perspective on the topic.
In case artificial intelligence tools were used, their way and purpose of usage has been made transparent.
Moreover, I have cited all my sources in accordance with academic standards.
I am ready and able to explain and defend the positions developed in this thesis.
This thesis has not been submitted, either in part or whole, at this or any other university.

\namesigdate
\pagenumbering{gobble}
\pagebreak


% --- Abstract ---
\begin{abstract}
  \textbf{\LARGE{Abstract}}\\\\
  Large language models are increasingly used as therapeutic agents in
  mental health applications, but it is unclear how consistently they
  respond when given the same patient input multiple times. This thesis
  measures therapeutic stability in the rescripting phase of Imagery
  Rehearsal Therapy, where the model must choose a therapeutic strategy
  and generate a response that carries it out.

  Ten models, including the EU-sovereign Mistral Large~3, two proprietary
  models (GPT-5.4, Claude Sonnet~4.6), and seven open-weight comparators,
  are evaluated across 6{,}000 trials at five temperature settings. A
  three-level evaluation framework compares strategy selections (Jaccard
  similarity), response texts (BERTScore), and whether the model does what
  it declared (LLM judge alignment).

  The main findings are: (1)~Mistral Large~3 matches the proprietary
  models on plan consistency at its recommended temperature range.
  (2)~Temperature affects which strategies the model picks but not how
  well it carries them out. The instability sits in strategy selection,
  not in response quality. (3)~Greedy decoding ($T = 0.0$) is not
  optimal for all models; three achieve higher consistency at $T = 0.075$.
  (4)~BERTScore captures whether the therapeutic tone stays similar but
  does not detect changes in the underlying strategy. The three levels
  together reveal patterns that no single metric shows on its own.

  At low temperatures, the EU-sovereign Mistral Large~3 reaches the same
  plan consistency as the proprietary models. At higher temperatures, the
  proprietary models remain stable while Mistral degrades. The framework
  itself is not specific to IRT and could be applied to other structured
  therapy protocols.
\end{abstract}


% --- Lists ---
\tableofcontents
\listoffigures
\listoftables


% --- Main matter ---
\pagenumbering{arabic}

\include{chapters/01_introduction}
\include{chapters/02_background}
\include{chapters/03_methodology}
\include{chapters/04_implementation}
\include{chapters/05_results}
\include{chapters/06_discussion}
\include{chapters/07_conclusion}
\include{chapters/07_acknowledgements}


% --- AI tools disclosure ---
\chapter*{Use of AI Tools in This Thesis}

Throughout the research and writing of this thesis,
I made use of several LLM-based tools in a supporting capacity.

For intellectual engagement with the subject matter,
I used Claude (Anthropic)\footnote{\href{https://claude.ai}{https://claude.ai}} and ChatGPT (OpenAI)\footnote{\href{https://chatgpt.com}{https://chatgpt.com}} as discussion partners.
Their role was to challenge my reasoning, surface counterarguments, and suggest alternative framings for the arguments I had already developed.
As English is not my native language, I occasionally used these tools to refine phrasing where my original formulations lacked precision or fluency.

For software development, I used Claude Code and Cursor (with Claude Sonnet) as coding assistants for the evaluation pipeline that forms the core of this thesis.
Given that the pipeline itself measures LLM behavioral stability, I took particular care to verify all AI-generated code before inclusion, to avoid introducing confounds into the instrument being studied.

No part of the thesis text was generated by AI.
All arguments, interpretations, and conclusions are my own.


% --- Glossary and bibliography ---
\glsaddall
\printglossaries
\printbibliography


% --- Appendix ---
\appendix
\include{chapters/appendix}

\end{document}
